---
title: "biostats_hw2"
output: github_document
date: "2023-10-06"
---

```{r}
library(tidyverse)
```


# Problem 1 
Suppose the probability of having at least one dental checkup during a two-year period is
73%. 56 random individuals are being studied for their health care utilization practices over a
two-year period. Compute “part a” by hand and show key steps. For all other parts you may
hand calculate or use R.
a) What is the probability that exactly 40 of these individuals will have at least one dental
checkup?
b) What is the probability that at least 40 of these individuals will have at least one dental
checkup?

## Part a
```{r}
exactly_40 = dbinom(40, 56, 0.73)
exactly_40
```
## Part b
```{r}
at_least_40 = 1 - pbinom(39, 56, 0.73)
at_least_40
```


# Problem 2
Suppose the number of tornadoes in the United States follows a Poisson distribution with
parameter 𝜆 = 6 tornadoes per year. Compute using tables or R. Show the formula for “part
a” (it can be handwritten and embedded in the pdf file).
a) What is the probability of having fewer than 3 tornadoes in the United States next year?
b) What is the probability of having exactly 3 tornadoes in the United States next year?
c) What is the probability of having more than 3 tornadoes in the United States next year?

## Part a
```{r}
fewer_than_3 = ppois(3, 6)
fewer_than_3
```
## Part b
```{r}
exactly_3 = dpois(3, 6)
exactly_3
```
## Part c
```{r}
more_than_3 = 1 - ppois(3, 6)
more_than_3
```


# Problem 3
Assume the systolic blood pressure of 20-29 year old American males is normally distributed
with population mean 128.0 and population standard deviation 10.2.
a) What is the probability that a randomly selected American male between 20 and 29
years old has a systolic blood pressure above 137.0?
b) What is the probability that the sample mean for blood pressure of 50 males between 20
and 29 years old will be less than 125.0?
c) What is the 90th percentile of the sampling distribution of the sample mean X for a
sample size of 40?

```{r}
mu = 128
sigma2 = 10.2
N = 10

set.seed(4)
  norm.df = tibble(
    sample = rnorm(N, mu, sigma2),
    id     = "Original Normal Distribution"
  )
  
norm.df |> 
    ggplot(aes(x = sample, y = ..density.., fill = id)) +
    geom_histogram(binwidth = 6, colour = "black", alpha = 0.1) +
    geom_density(alpha = 0.4) +
    labs(
      x = sprintf("Xi ~ N(%e, %f)", mu, sigma2),
      y = "Systolic Blood Pressure",
      title = "Histogram and Density Curve for Systolic Blood Pressure for 
      American Males 20 to 29 years old") +
 scale_fill_viridis_d("") +
theme_classic() +
theme(legend.position = "none")
```
## Part a
```{r}
bp_above_137 = 1-pnorm(137, 128, 10.2)
bp_above_137
```
## Part b
```{r}
sd = 10.2/(sqrt(50))
bp_under_125 = pnorm(125, 128, sd)
bp_under_125
```
## Part c
```{r}
std = 10.2/(sqrt(40))
c_90 = qnorm(.9, 128, std)
c_90
```


# Problem 4
Some researchers are interested in the mean pulse of young women suffering from fibromyalgia.
They selected a random sample of 40 young females suffering from fibromyalgia. The sample
mean of their pulses was 80 and the sample standard deviation was 10.
a) Compute the 95% confidence interval for the population mean pulse rate of young females
suffering from fibromyalgia.
c) Suppose the researchers now want to test the null hypothesis that the mean pulse of
young women suffering from fibromyalgia is equal to 70, against the alternative that the
mean pulse is not equal to 70, at the 𝛼 = 0.01 significance level. Conduct this hypothesis
test, and interpret the results.

## Part a
```{r}
lower = 80 - 2.021 * 10/sqrt(40)
upper = 80 + 2.021 * 10/sqrt(40)
CI = c(lower, upper)
CI
```
## Part c
```{r}
tstat = (80 - 70) / (10/sqrt(40))
tstat
```


  
  