---
title: "biostats_hw3"
output: github_document
date: "2023-10-28"
---

```{r}
library(tidyverse)
library(dplyr)
library(MASS)
library(Hmisc)
library(gt)
library(gtsummary)
```


## Question 1 
Some medical professionals claim that the average weight of American women is 171 pounds.
The column lwt holds the mother’s weight (in pounds) at last menstrual period, i.e. her
pre-pregnancy weight. Use this column for the following questions.
a) Construct a 95% confidence interval of true mean weight of American women.
b) Interpret the confidence interval.
c) Comment on the validity of the statement above (“Some medical professionals claim that
the average weight of American women is 171 pounds”). In other words, what can we
say about this statement given our confidence interval from part a?

## Part a
```{r}
CI = t.test(birthwt$lwt, conf.level = 0.95)$conf.int
CI
```

## Part b
From our calculated confidence interval, we can say that we are 95% confident that the average weight of women will be between 125.4270 and 134.2027 lbs.

## Part c
The statement "some medical professionals claim that the average weight of American women is 171 pounds" is a false claim because we have calculated frmo the data that the average weight of women is 95% likely to be between 125.4270 and 134.2027 lbs which does not include 171 lbs.

## Question 2
In this data set, we have a variable (smoke) indicating the smoking status of the mothers
during pregnancy. Some doctors believe that smoking status is related to weight. Using the
columns smoke and lwt, test this claim. (Note: a value of 1 indicates the mother is in the
“smoking” group.)
a) Test for the equality of variances between the two groups. (Use a 5% significance level.)
b) Given your answer from part a, what kind of hypothesis test will you perform?
c) Conduct your chosen hypothesis test from part b at the 10% significance level. What is
your decision regarding the null? Interpret this result in the context of the problem.

## Part a
```{r}
smoking = birthwt$lwt[birthwt$smoke == 1]
non_smoking = birthwt$lwt[birthwt$smoke == 0]
var = var.test(smoking, non_smoking)
var
```

Thus, we fail to reject the null hypothesis and conclude that the variances are not statistically different. 

## Part b
We found that the variances are not statistically different, so we will conduct a two-sample
t-test for equal variances by pooling the standard deviations.

## Part c
```{r}
n1 = sum(birthwt$smoke == 1)

sd1 = sd(birthwt$lwt[which(birthwt$smoke == 1)])

n2 = sum(birthwt$smoke == 0)

sd2 = sd(birthwt$lwt[which(birthwt$smoke == 0)])

pooled_sd = sqrt(((n1 - 1) * sd1^2 + (n2 - 1) * sd2^2)/(n1 + n2 - 2))

sample_mean_1 = mean(birthwt$lwt[which(birthwt$smoke == 1)])

sample_mean_2 = mean(birthwt$lwt[which(birthwt$smoke == 0)])

t_stat = (sample_mean_1 - sample_mean_2) / (pooled_sd * sqrt(1/n1 + 1/n2))

t_cv = qt(1 - 0.10/2, df = n1 + n2 - 2)

glue::glue("mean weight in smoking group = {round(sample_mean_1, 2)}
mean weight in non-smoking group = {round(sample_mean_2, 2)}
pooled standard deviation = {round(pooled_sd, 2)}
t test statistic = {round(t_stat, 2)}
t critical value = {round(t_cv, 2)}")
```

Thus, at the 10% significance level, we fail to reject the null hypothesis. We do not have enough evidence to prove that the mean weight of mothers is different between the smoking and the non-smoking groups. 

## Question 3
According to the CDC, approximately 20% of pregnant American women suffer from hyper-
tension. Do our data support this claim? (Use column ht - a value of 1 means the mother is
suffering from hypertension.)
a) Conduct a 99% confidence interval and interpret the results. What can we conclude
about the CDC’s claim from this interval?
b) Conduct a one-sided hypothesis test at the 𝛼 = 0.1 level. In this test, we want to see if
the true proportion is indeed less than the claimed 20%. What can we conclude about
the CDC’s claim?

## Part a
```{r}
smp_prp = sum(birthwt$ht == 1) / nrow(birthwt)
alpha = 0.01
z = qnorm(1 - alpha/2)
error = z*sqrt((smp_prp*(1 - smp_prp)) / nrow(birthwt))
CI = c(smp_prp - error, smp_prp + error)
CI
```

The true proportion of pregnant women who suffer from hypertension is between 0.02780412 and 0.10918001.

## Part b
```{r}
alpha = 0.1
z = qnorm(alpha)
std_error = sqrt((smp_prp * (1 - smp_prp)) / nrow(birthwt))
t_stat = (smp_prp - 0.2) / std_error
if (t_stat < z) {
  cat("reject null: there is sufficient evidence to support the claim")
} else {
  cat("fail to reject null: there is insufficient evidence to support the claim")
}
```

Since our p value is small enough and less than our alpha-level, we are able to reject the null hypothesis. 

## Question 4
Is there a difference between uterine irritability in the group of pregnant women who smoke
vs the group of pregnant women that don’t smoke? (Use columns ui and smoke.)
Conduct a hypothesis test at the 𝛼 = 0.01 level. What can we conclude about the proportions
of women with uterine irritability between the smoking groups?

```{r}
birthwt |> 
gtsummary::tbl_cross(row = smoke, col = ui, label = list(smoke ~ "Smoking Status", ui ~ "Uterine Irritability")) |> 
gtsummary::bold_labels() |> 
gtsummary::as_gt() |> 
gt::as_latex()


x1 = 13 
n1 = 74 
p1 = x1/n1


x2 = 15 
n2 = 115 
p2 = x2/n2

prop.test(c(x1, x2), c(n1, n2),
alternative = "two.sided",
conf.level = 0.99)
```

At the 1% significance level, the p-value is 0.5191. This value is greater than the alpha. Thus, we fail to reject the null hypothesis.

## Question 5
Is race related to birth weight? (Use columns race and bwt.)
a) What test would be most appropriate to answer this question?
b) What assumptions are we making by using this test? 
c) Conduct the test at the 5% significance level and interpret your results. Be sure to write
the hypotheses you are testing.

## Part a
The most appropriate test for this question is an ANOVA test because we are comparing multiple groups. 

## Part b
The assumptions of an ANOVA test are:
1. there are k populations of interets
2. the samples are drawn independently from the underlying populations
3. there is homoscedasticity and normality

## Part c
The null hypothesis H0: there is no difference between birth weight between different racial groups. 
The alternate hypothesis H1: at least one racial group has different birth weight compared to the other racial group. 

```{r}
anova = aov(bwt ~ race, data = birthwt)
summary(anova)
```

Results show a p-value of 0.008. Thus we can reject the null hypothesis. We have evidence that at least
two of the means are not equal. 